Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2025-02-22 analysis/paper/paper.qmd
work on the result section

To construct a reliable fine-grained chronological sequence, archaeological deposits were divided into six temporal layers spanning from the 14th century to the 19th century based on excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators. These diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks [@Wang2011; @Hsieh2009]. Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above could offer a relatively robust framework for comparing archaeological records before and after the European influence. Among these six temporal layers, the fourth layer can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second layer corresponds to the 19th century, associated with the large influx of Chinese migrants in Yilan.

Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences. Deer antlers that were not attached to cranial bones were excluded since they might not directly related to dietary practices. To obtain generalized pattern of resource use, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010]. Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units per cultural layers. Taxonomic abundance was quantified using the Number of Identified Specimens (NISP) method [@Grayson2004; @Lyman2012]. Different from the minimum number of individuals (MNI) that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of dietary practices by considering all fragmented bone assemblages. Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site with less impacts from sample size, and therefore is used throughout this study [@Grayson1984; @Lyman2008]. For comparison of identified taxa across classes, the normalized NISP were used to avoid potential biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004]. The normalized NISP is calculated by dividing NISP by the average number of skeletal elements of that given taxa.

Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the frequency of skeletal elements and surface modifications such as cutmarks of deer bones were examined to assess butchery and transport strategy. This can also shed light on how indigenous people utilize deer in response to high economic demand when involved in regional trade network. Skeletal elements was quantified using NISP and classified into groups based on meat yield: meaty (e.g., upper forelimb and upper hindlimb ), mid-meaty (e.g., body parts), and less meaty portions (e.g., skull and lower limb) [@Madrigal2004; @Munro2004]. Teeth without associated cranial elements were excluded here to reduce overrepresentation due to their higher preservation compared to othe parts. While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain [@Madrigal2002]. Thus, less meaty elements were considered as secondary source of nutrition in the interpretation. Similarly, the degree of fragmentation could also imply the extraction of marrow or grease. The degree of fragmentation is calculated by dividing Minimal Number of Elements (MNE) by NISP. For butcher patterns, cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were identified and quantified for a comparison between different temporal sequences. In addition, modifications such as cut marks, fracturing, and burning were examined, especially the cut marks associated with the skinning process. Skinning processes can be identified by the encircling and straight marks left on the lower limb, such as metapodials, which result from the series of cuts typically made to acquire intact animal hides [@Costamagno2019; @Isaakidou2007].

@@

fauna_combined_context_all <-
  rbind(fauna_sam, fauna_H_sam) %>% # combine the two datasets
  filter(!`部位/名稱` %in% c("角","犄角","角?")) %>%  #not associated with diet
  filter(!`部位/名稱`== "角基部"|!is.na(`部位/左右`)) %>% # associated with cranial parts
  mutate(`重量(g)` = as.numeric(`重量(g)`)) %>%
  group_by(taxa) %>%
  mutate(`Weight (g)` = sum(`重量(g)`, na.rm = T)) %>%
  ungroup() %>%
  mutate(period = factor(period, levels = c("CL1","CL2","CL3","CL4","CL5","CL6"), order = T)) %>%
  mutate(category = case_when(str_detect(animal, "sika")|str_detect(animal, "sambar")|
                              str_detect(animal, "e deer")~ "deer", str_detect(animal, "fish") ~ "fish",
                              str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "cattle",
                              animal == "aves" ~ "bird", TRUE ~ animal)) %>%
  mutate(class = case_when(str_detect(taxa, "muntjac")|str_detect(taxa, "ammal")|str_detect(taxa, "erv")|
                           str_detect(taxa, "Sus")|str_detect(taxa, "Rattus")|str_detect(taxa, "Rusa")|
                           str_detect(taxa, "Bos")|str_detect(taxa, "Bubalus")~ "mammal",
                           str_detect(category, "fish") ~ "fish", str_detect(category, "bird") ~ "bird",
                           TRUE ~ "reptile")) %>%
  mutate(`部位/左右` = case_when(`部位/左右` == "Ｌ" ~ "L", `部位/左右` == "Ｒ" ~ "R", TRUE ~ `部位/左右`)) %>%
  mutate(category = fct_relevel(category, "deer","boar","muntjac","cattle","rat","bird","fish","turtle")) %>%  mutate(modify = case_when(str_detect(`人為痕跡`, "切")|str_detect(`人為痕跡`, "削")|
                            str_detect(`人為痕跡`, "砍")~ "cutmarks",
                            str_detect(`人為痕跡`, "火")|str_detect(`人為痕跡`, "燒")~ "burn", TRUE ~ "no")) %>%filter(is.na(refitted)) # for NISP

@@

```{r}
#| label: taxa-NISP-counts
#| echo: false
# taxa counts
fauna_combined_taxa <-
  fauna_combined_context_all %>%
  count(taxa, `Weight (g)`) %>%
  mutate(taxa = str_replace(taxa, "cervidae", "cervid")) %>%
  mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Unidentified", tolower(taxa), sep =" "),
                          TRUE ~ taxa)) %>%
  mutate(`Common name` = case_when(str_detect(taxa, "Cer") ~ "Sika deer", str_detect(taxa, "Bub") ~ 
                                   "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",   
                                   str_detect(taxa, "Ree") ~ "Muntjac", str_detect(taxa, "Sus") ~ "Boar", 
                                   str_detect(taxa, "Bos") ~ "Yellow cattle", str_detect(taxa, "Rat") ~ 
                                   "Rat", str_detect(taxa, "Pli") ~ "Catfish", str_detect(taxa, "Act") ~ 
                                   "Ray-finned fish", str_detect(taxa, "Tes") ~ "Turtle",
                                   str_detect(taxa, "cervid") ~ "Deer family")) %>%
  mutate(group = case_when(str_detect(`Common name`, "deer") ~ 1, str_detect(`Common name`, "Deer") ~ 2,
                           str_detect(taxa, "Sus") ~ 3, str_detect(taxa, "Bub")|str_detect(taxa, "Bos") ~ 4,
                           str_detect(taxa, "munt") ~ 5, str_detect(taxa, "Rat") ~ 6, 
                           str_detect(`Common name`, "fish") ~ 7, str_detect(taxa, "Tes") ~ 8,
                           str_detect(taxa, "Ave") ~ 9, str_detect(taxa, "large m") ~ 10,
                           str_detect(taxa, "medium m") ~ 11, TRUE ~ 12)) %>%
  group_by(group) %>%
  arrange(desc(`Weight (g)`), .by_group = TRUE) %>%
  ungroup () %>%
  rename(Taxon = taxa, NISP = n) %>%
  select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
  bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>% 
  mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`))

total_NISP <- fauna_combined_taxa %>% select(NISP) %>% tail(1) #rev(fauna_combined_taxa$NISP)[1]
unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Uni")) %>% summarise(sum(NISP))
deer_NISP <- fauna_combined_taxa %>% 
  filter(str_detect(`Common name`, "eer")|str_detect(Taxon, "Ree")) %>% summarise(sum(NISP))
deer_NISP_per <- round(deer_NISP/(total_NISP-unident), 3)*100
```

The NISP of faunal assemblages from Kiwulan analyzed in this study is `r total_NISP`, where deer family is the majority that accounts for about `r deer_NISP_per`% of the identified bones, followed by boar and cattle as the second and third taxon (@tbl-counts). Within the assemblages from the deer family, sika deer is the most common animal, with a few assemblages identified as muntjac and a specimen indicating a possible affiliation with sambar deer. Although sika deer was the major animal source at the site, variations in animal use could be observed between different cultural layers across 600 years. To highlight general patterns of dietary practices in terms of their economic values over time, the deer family were separated into two categories based on body size, deer (sika deer, sambar deer, and large deer) and muntjac, while water buffalo and yellow cattle were combined into one category, cattle. Rats (Rattus sp.) were excluded from further analysis as they are normally considered commensal animals, and their relatively intact skeletons from one single archaeological context show no direct evidence of being part of the daily diet.

@@

For animal classes, the relative abundance shows that mammals are the primary dietary source throughout all layers, with fish as the secondary source in CL4 and birds in CL3 and CL2 (@fig-NISP-proportion: A). Fish and birds are both identified from CL5 to CL3; however, their proportions (less than 10%) are much lower compared to mammals. The reptile is represented by one piece of turtle shell found in CL4, which might not be a common food content at the site. Within those identifiable mammals, deer is the major source from CL6 to CL2 that accounts for at least 85% per layer, while CL1 is predominated only by cattle (@fig-NISP-proportion: B). Wild boar was present from CL6 to CL3, and had a higher proportion in CL4 (`r CL4_boar` %), while muntjac appeared in CL5, and was continuously present until CL2. In general, there is an increase in animal diversity in CL4 and CL5, especially in CL4, where boar and fish have slightly higher proportions compared with other layers. Given that the taxonomic richness, or animal diversity, could be impacted by sample size varied from layer to layer, the relationship between the number of taxa and sample size (NISP) was examined by linear regression model. The result shows no correlation between the diversity of animals and sample size ($R^2$ = `r div_lm[1]`, P = `r div_lm[2]`), which rules out the potential sample size bias.

```{r}
#| label: fig-NISP-proportion
#| echo: false
#| fig-cap: "A: Relative abundance of vertebrate classes by cultural layers based on normed NISP. The data is derived from Table 1. B: Relative abundance of mammals by cultural layers based on NISP. CL6 to CL1 prepresent cultural layers from the earliest to the latest layer. The NISP for each layer is CL6: 46, CL5: 65, CL4: 156, CL3: 28, CL2: 12, and CL1: 25."

@@

Since deer remain was the major component of faunal assemblage at Kiwulan according to their high quantity and ubiquity, the frequency of skeletal elements was examined to explore any changes in processing decisions of meat portions between different phases. Deer bones were grouped into five body portions based on their nutritional value [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), and lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus, and phalanges). The result (@fig-deer-portion: A) shows that the meaty portion, including upper forelimb and upper hindlimb, especially upper forelimb, is the most frequent part throughout the cultural layers, followed by the less-meaty portions, such as lower limb and foot. The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together (e.g., @Stiner1991), or marrows from upper limb were commonly utilized. It should be noted that the frequencies of axial elements, such as ribs, were generally low across different phases, likely due to a high degree of fragmentation, butchery practices, or dietary preferences. A different pattern is observed in the frequency of head bones, which were a few in CL6, increased to the highest number in CL4, and then decreased thereafter.

Considering varying frequencies of skeletal parts might be mediated by differential bone density, the relationship between the survivorship of deer skeletal elements and bone mineral density was tested by Pearson correlation. The values of bone survivorship were collected from @Lyman1994's research on deer, and calculated by dividing NISP by expected NISP, followed by multiplying 100%. The result (@fig-deer-portion: B) shows that there is no significant positive correlation ($r$ = `r survivor_cor[1]`, P = `r survivor_cor[2]`)) between skeletal part survivorship and bone density, suggesting that density-mediated bias has little impact on the taphonomic processes of deer bones.

@@

@@

Other than quantity, the measure of presence and absence of a given taxon across the sampeld units could give insights into the frequency and distribution consistency of taxa at a settlement scale.

@@

>>>>

To construct a reliable fine-grained chronological sequence, archaeological deposits were divided into six cultural layers (CL6-CL1) spanning from the 14th century to the 19th century based on excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators. These diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks [@Wang2011; @Hsieh2009]. Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above could offer a relatively robust framework for comparing archaeological records before and after the European influence. Among these six temporal layers, the fourth layer can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second layer corresponds to the 19th century, associated with the large influx of Chinese migrants in Yilan.

Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences. Deer antlers that were not attached to cranial bones were excluded since they might not directly related to dietary practices. To obtain generalized pattern of resource use, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010]. Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units per cultural layers. Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when necessary, such as for estimating the expected NISP at a site [@Grayson2004; @Lyman2012]. Different from the minimum number of individuals (MNI) that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of dietary practices by considering all fragmented bone assemblages. Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site with less impacts from sample size [@Grayson1984; @Lyman2008]. For comparison of identified taxa across classes, the normalized NISP were used to avoid potential biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004]. The normalized NISP is calculated by dividing NISP by the number of skeletal elements.

Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the frequencies of skeletal elements and surface modifications of deer bones were examined to assess butchery and transport strategies. This can shed light on how indigenous people utilize deer in response to high economic demand when involved in regional trade network. Skeletal elements were quantified using NISP and classified into groups based on meat yield: meaty (e.g., upper forelimb and upper hindlimb), mid-meaty (e.g., body parts), and less meaty portions (e.g., skull and lower limb) [@Madrigal2004; @Munro2004]. Teeth without associated cranial elements were excluded here to reduce overrepresentation due to their higher preservation compared to other parts. While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain [@Madrigal2002]. Thus, less meaty elements were considered as secondary source of nutrition in the interpretation. Similarly, the degree of fragmentation could also imply the extraction of marrow or grease, which is calculated by dividing the Minimal Number of Elements (MNE) by NISP. Bone surface modifications relative to human butchering and cooking activities, such as cut marks, fracturing, and burning damage, were recorded, as well as natural marks such as weathering and root etching for understanding taphonomic effects. Cut marks on major limb joints, such as shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were further quantified to explore any changes in butchering patterns between temporal sequences. Moreover, cut marks on lower limbs relative to skinning processes were examined and compared. Those marks are charactersitic of circular and straight incisions left by the series of cuts for acquiring intact animal hides [@Costamagno2019; @Isaakidou2007].

@@

fauna_combined_context_all <-
  rbind(fauna_sam, fauna_H_sam) %>% # combine the two datasets
  filter(!`部位/名稱` %in% c("角","犄角","角?")) %>%  #not associated with diet
  filter(!`部位/名稱`== "角基部"|!is.na(`部位/左右`)) %>% # associated with cranial parts
  mutate(`重量(g)` = as.numeric(`重量(g)`)) %>%
  group_by(taxa) %>%
  mutate(`Weight (g)` = sum(`重量(g)`, na.rm = T)) %>%
  ungroup() %>%
  mutate(period = factor(period, levels = c("CL1","CL2","CL3","CL4","CL5","CL6"), order = T)) %>%
  mutate(category = case_when(str_detect(animal, "sika")|str_detect(animal, "sambar")|
                              str_detect(animal, "e deer")~ "deer", str_detect(animal, "fish") ~ "fish",
                              str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "cattle",
                              animal == "aves" ~ "bird", TRUE ~ animal)) %>%
  mutate(class = case_when(str_detect(taxa, "muntjac")|str_detect(taxa, "ammal")|str_detect(taxa, "erv")|
                           str_detect(taxa, "Sus")|str_detect(taxa, "Rattus")|str_detect(taxa, "Rusa")|
                           str_detect(taxa, "Bos")|str_detect(taxa, "Bubalus")~ "mammal",
                           str_detect(category, "fish") ~ "fish", str_detect(category, "bird") ~ "bird",
                           TRUE ~ "reptile")) %>%
  mutate(`部位/左右` = case_when(`部位/左右` == "Ｌ" ~ "L", `部位/左右` == "Ｒ" ~ "R", TRUE ~ `部位/左右`)) %>%
  mutate(category = fct_relevel(category, "deer","boar","muntjac","cattle","rat","bird","fish","turtle")) %>%
  mutate(cutmarks = ifelse(str_detect(`人為痕跡`, "切"),"yes", "no"),
         fractures = ifelse(str_detect(`人為痕跡`, "削")|str_detect(`人為痕跡`, "砍")|
                            str_detect(`人為痕跡`, "折"), "yes", "no"),
         burning =  ifelse(str_detect(`人為痕跡`, "火")|str_detect(`人為痕跡`, "燒"), "yes", "no")) %>%
  filter(is.na(refitted)) # for NISP

@@

```{r}
#| label: taxa-NISP-counts
#| echo: false
# taxa counts
fauna_combined_taxa <-
  fauna_combined_context_all %>%
  count(taxa, `Weight (g)`) %>%
  mutate(taxa = str_replace(taxa, "cervidae", "cervid")) %>%
  mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Indeterminate", tolower(taxa), sep =" "),
                          TRUE ~ taxa)) %>%
  mutate(`Common name` = case_when(str_detect(taxa, "Cer") ~ "Sika deer", str_detect(taxa, "Bub") ~ 
                                   "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",   
                                   str_detect(taxa, "Ree") ~ "Muntjac", str_detect(taxa, "Sus") ~ "Boar", 
                                   str_detect(taxa, "Bos") ~ "Yellow cattle", str_detect(taxa, "Rat") ~ 
                                   "Rat", str_detect(taxa, "Pli") ~ "Catfish", str_detect(taxa, "Act") ~ 
                                   "Ray-finned fish", str_detect(taxa, "Tes") ~ "Turtle",
                                   str_detect(taxa, "cervid") ~ "Deer family")) %>%
  mutate(group = case_when(str_detect(`Common name`, "deer") ~ 1, str_detect(`Common name`, "Deer") ~ 2,
                           str_detect(taxa, "Sus") ~ 3, str_detect(taxa, "Bub")|str_detect(taxa, "Bos") ~ 4,
                           str_detect(taxa, "munt") ~ 5, str_detect(taxa, "Rat") ~ 6, 
                           str_detect(`Common name`, "fish") ~ 7, str_detect(taxa, "Tes") ~ 8,
                           str_detect(taxa, "Ave") ~ 9, str_detect(taxa, "large m") ~ 10,
                           str_detect(taxa, "medium m") ~ 11, TRUE ~ 12)) %>%
  group_by(group) %>%
  arrange(desc(`Weight (g)`), .by_group = TRUE) %>%
  ungroup () %>%
  rename(Taxon = taxa, NISP = n) %>%
  select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
  bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>% 
  mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`))

total_NISP <- fauna_combined_taxa %>% select(NISP) %>% tail(1) #rev(fauna_combined_taxa$NISP)[1]
unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Uni")) %>% summarise(sum(NISP))
deer_fami_NISP <- fauna_combined_taxa %>% 
  filter(str_detect(`Common name`, "eer")|str_detect(Taxon, "Ree")) %>% summarise(sum(NISP))
deer_fami_NISP_per <- round(deer_fami_NISP/(total_NISP-unident), 3)*100
```

The faunal assemblage analyszed in this paper from Kiwulan is composed of `r total_NISP` bone fragments based on NISP, which increased over time from CL6, peaked at CL4, and then declined dramatically after CL4 (@fig-NISP-proportion: A). A similar trend also observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B). Among the funal remains, deer family is the majority that accounts for about `r deer_fami_NISP_per`% of the identified bones, followed by boar and cattle as the second and third taxon (@tbl-counts). Within the deer family, sika deer is prevalent, with a few bones identified as muntjac and a specimen indicating a possible affiliation with sambar deer. Although sika deer was the major animal source at the site, variations in their quantities and proportions could be observed between temperoal sequences across 600 years. To further compare general patterns of dietary practices in terms of their economic values over time, the deer family were separated into two categories based on body size, deer (sika deer, sambar deer, and large deer) and muntjac, while water buffalo and yellow cattle were combined into one category, cattle. Rats (Rattus sp.) were excluded from further analyses as they are normally considered commensal animals, and their relatively intact skeletons from one single archaeological context show no direct evidence of being part of the daily diet.

@@

For animal classes, the relative abundance shows that mammals are the primary dietary source throughout all layers, with fish as the secondary source in CL4 and birds in CL3 and CL2 (@fig-NISP-proportion: C). Fish and birds are both identified from CL5 to CL3; however, their proportions (less than 10%) are much lower compared to mammals. The reptile is represented by one piece of turtle shell found in CL4, which might not be a common food content at the site. Within those identifiable mammals, deer is the major source from CL6 to CL2 that accounts for at least 85% per layer, while CL1 is predominated only by cattle (@fig-NISP-proportion: D). Wild boar was present from CL6 to CL3, and had a higher proportion in CL4 (`r CL4_boar` %), while muntjac appeared in CL5, and was continuously present until CL2. In general, there is an increase in animal diversity in CL4 and CL5, especially in CL4, where boar and fish have slightly higher proportions compared to those in other layers. Given that the taxonomic richness, or animal diversity, could be impacted by sample size varied from layer to layer, the relationship between the number of taxa and sample size (NISP) was examined by linear regression model. The result shows no correlation between the diversity of animals and sample size ($R^2$ = `r div_lm[1]`, P = `r div_lm[2]`), which rules out the potential sample size bias.

```{r}
#| label: fig-NISP-proportion
#| echo: false
#| fig-cap: "A: The number of identified specimen per cultural layers. The NISP data is derived from Table 1 . CL6 represents the earliest layer and CL1 is the youngest. B: The weight of the unidentified small bone fragments for each cultural layer. C: Relative abundance of vertebrate classes by cultural layers based on normed NISP. D: Relative abundance of identified mammals by cultural layers based on NISP."

@@

Since deer is the most frequent animal at Kiwulan, the frequency of skeletal parts was examined to explore any changes in processing decisions of meat portions between different phases. Deer bones were grouped into five body portions based on their nutritional value [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), and lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus, and phalanges). The result (@fig-deer-portion: A) shows that the meaty portion, especially the upper forelimb, is the most frequent part throughout the cultural layers, followed by upper hindlimb and the less-meaty portions, such as lower limb and foot. The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together (e.g., @Stiner1991). The frequencies of axial elements, such as ribs, were generally low at Kiwulan, likely due to a high degree of fragmentation, butchery practices, or dietary preferences. A different pattern between layers is observed in the frequency of head bones, which were a few in CL6, increased to the highest number in CL4, and then decreased thereafter. In addition, lower limb and foot bones appear to be proportionally less in CL2 and CL3 compared to earlier periods.

Considering the impacted of differential bone density on preservation of skeletal elements, the relationship between the survivorship of deer skeletal elements and bone mineral density was tested by Pearson correlation. The values of bone survivorship were collected from @Lyman1994's research on deer, and calculated by dividing NISP by expected NISP, followed by multiplying 100%. The result (@fig-deer-portion: B) shows that there is no significant positive correlation ($r$ = `r survivor_cor[1]`, P = `r survivor_cor[2]`)) between bone survivorship and bone density, suggesting that density-mediated bias has little impact on the taphonomic processes of deer bones.

@@

```{r}
#| label: tbl-deer-modify
#| echo: false
#| tbl-cap: "Number and proportion of deer bones with surface modifications at Kiwulan."
#| tbl-cap-location: bottom

fauna_deer_only <-
  fauna_combined_context %>%
  filter(category %in% c("deer") & !is.na(period))

deer_NISP <- fauna_deer_only %>% count(period) %>% rename(NISP = n)

# bones with surface modifications
burned_deer <- fauna_deer_only %>% filter(burning == "yes") %>% count(period) %>% rename(burned = n)
cutmark_deer <- fauna_deer_only %>% filter(cutmarks == "yes") %>% count(period) %>% rename(cutmarks = n)
frac_deer <- fauna_deer_only %>% filter(fractures  == "yes") %>%
  filter(`部位/名稱` %in% c("脛骨","股骨","肱骨","橈骨","上肢骨","掌骨","掌骨或蹠骨","蹠骨")) %>%
  count(period) %>% rename(fractures = n)
list_modif <- list(burned_deer, cutmark_deer, frac_deer, deer_NISP)

# join
deer_modify <-
  list_modif %>%
  reduce(inner_join, by = 'period') %>%
  mutate(cut_per = round(cutmarks/NISP, 3) * 100, burn_per = round(burned/NISP, 3) * 100,
         frac_per = round(fractures/NISP, 3) * 100) %>%
  mutate(burned = paste(burned, paste0("(", burn_per, " %)")),
         cutmarks = paste(cutmarks, paste0("(", cut_per, " %)")),
         fractures = paste(fractures, paste0("(", frac_per, " %)"))) %>%
  rename(`cut marks` = cutmarks) %>%
  select(-cut_per, -frac_per, -burn_per)

knitr::kable(deer_modify)
```

Surface modifications, including cut marks, fractures, and burning damage, were identified on deer bones from CL6 to CL2, with variations in their proportions (@tbl-deer-modify). Burned bones, characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4. Cut marks were present on 50% of bones in CL2, 30% of bones in earlier period (CL6 to CL4), and at a lower proportion of 23% in CL3. Most cut marks were located on major limb joints and long bones, suggesting butchering activities related to disarticulation and meat detaching. @fig-deer-cutmark shows that at least 50 % of joints of each deer part in CL4 present cut marks, with a higher proportion of 75% at the hip. A similar pattern was observed in CL5, but with a lower proportion of cut marks on the ankle, knee, and hip compared to CL4. Not all joints were found in CL6, CL3 and CL2, making direct comparisons across all layers difficult. Nevertheless, the overall high proprotion of joints with cut marks suggests intensive proceessing of deer throughout the site's occupation. Cut marks on lower limbs can be associeted with skinning and were primarily found on metapodials in CL6 (n = 1), CL5 (n = 4), CL4 (n = 6), CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1).

Bone breakage patterns, including spiral fractures, transverse fractures, longitudinal splits, percurssion scars, and chop marks [@Haynes1983; @Gifford2018; @Fisher1995], were observed esepically on long bones across different layers. These fractures indicate intentional marrow extraction at Kiwulan. CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5. This suggests more intensive marrow extraction in later periods. Most chop and cut marks were likly made by metal tools according to clean, smooth cut surfaces and finer marks on bones. The use of metal tools can be further supported by the finding of iron knives at Kiwulan [@Chen2007]. In addition to anthropogenic modifications, natural alterations to bones resulting from post-deposition impacts, such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@tbl-deer-modify).

```{r}
#| label: fig-deer-cutmark
#| echo: false
#| fig-cap: "The proportion of cut marks on major limb joints of deer across cultural layers. The method for clasifying and quantifiying limb joints is based on Lyman (2005)."

knitr::include_graphics(here::here("analysis/figures/deer_cutmarks.png"))
```

@@

frequencies of deer bone elements are consistent with the proportion of joints. 
Other than quantity, the measure of presence and absence of a given taxon across the sampeld units could give insights into the frequency and distribution consistency of taxa at a settlement scale.

@@

These post-depositional impacts might have led to an underestimation of cut marks due to the diffculty of surface identification.
